---
title: "Out of house data part1: prepeare for machine learning"
author: "Mehmet Umut Caglar"
date: "9/25/2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```



```{r initials, include=FALSE}
###*****************************
# INITIAL COMMANDS TO RESET THE SYSTEM
rm(list = ls())
if (is.integer(dev.list())){dev.off()}
cat("\014")
###*****************************

###*****************************
#Load Libraries
require("tidyverse")
require("DESeq2")
```

## The generation process for out of house data (Part 1/2)

The out of house protein data is taken from:
Schmidt, Alexander, et al. "The quantitative and condition-dependent 
Escherichia coli proteome." Nature biotechnology 34.1 (2016): 104.

The data set composes of two distinct sets; dataset 1 and data set2. They also include a combined data set.

```{r Getting the data}
dataSet1_notClean_notNormalized = read.csv(file = "SuppFor_Paper2/tableS4_PCount.csv")
dataSet2_notClean_notNormalized = read.csv(file = "SuppFor_Paper2/tableS5_PCount.csv")
dataSet_comb_notClean_notNormalized = read.csv(file = "SuppFor_Paper2/tableS6_PCount.csv")
```

Also want to check is combined data is coming from data set 1 and data set 2

```{r Comparing data sets}
###*****************************
# Check if combined is coming from 1 and 2
dataSet_comb_notClean_notNormalized %>%
  dplyr::filter(Dataset == 1) %>%
  dplyr::select(Gene, Glucose) %>%
  dplyr::arrange(Gene) %>%
  head(.)

dataSet1_notClean_notNormalized %>%
  dplyr::select(Gene, glucose) %>%
  dplyr::arrange(Gene) %>%
  head(.)

dataSet_comb_notClean_notNormalized %>%
  dplyr::filter(Dataset == 2) %>%
  dplyr::select(Gene, Glucose) %>%
  dplyr::arrange(Gene) %>%
  head(.)

dataSet2_notClean_notNormalized %>%
  dplyr::select(Gene, Glucose) %>%
  dplyr::arrange(Gene) %>%
  head(.)
```
it seems;
the data set 2 and combined data with data set == 2 
are consistent but,
the data set 1 and combined data with data set == 1 are NOT consistent 


I will go with both dataSet 1 and data set 2



Now we do not need all the samples in data set 1 & 2; the matching samples are:
* Glucose (in exponential phase)
* Glycerol (in exponential phase)
* Osmotic.stress.glucose (exponential phase glicose with 50 mm NaCl)
* Stationary.phase.1.day (Glucose in stationary phase)
* Stationary.phase.3.days (Glucose in late stationary phase)


```{r Cleaning data set 1 & 2}
# Selecting necessary columns for dataSet1
dataSet1_notClean_notNormalized %>%
  dplyr::select(gene_name = Gene, 
                glucose = glucose,
                glycerol = glycerol,
                Na_50_stress = X50.mM.NaCl,
                stationary = stationary.1.day,
                late_stationary = stationary.3.days)%>%
  dplyr::filter(gene_name!="")%>% # there are two rows with gene name ""
  dplyr::arrange(gene_name)-> dataSet1_notNormalized

head(dataSet1_notNormalized)

# Selecting necessary columns for dataSet2
dataSet2_notClean_notNormalized %>%
  dplyr::select(gene_name = Gene, 
                glucose = Glucose,
                glycerol = Glycerol,
                Na_50_stress = Osmotic.stress.glucose,
                stationary = Stationary.phase.1.day,
                late_stationary = Stationary.phase.3.days)%>%
  dplyr::filter(gene_name!="")%>% # there are two rows with gene name ""
  dplyr::arrange(gene_name)-> dataSet2_notNormalized

head(dataSet2_notNormalized)
```

We need to match what we measure and what those datasets have;

To do this need to update the `gene_name` with `Protein_id`'s (YP_ names) and generate the meta data table. For this we need the protein dictionary


```{r Replacing gene names with YP_ names in data set 2 }
dictionary = read.csv("nameDictionary_RNA&Protein.csv")
head(dictionary)

# Making changes for dataSet1
# ***********************************************

# combine dictionary and datatSet1
dictionary %>%
  dplyr::select(gene_name, Protein_id) %>%
  dplyr::left_join(dataSet1_notNormalized, .) -> dataSet1_notNormalized

# get rid of not matching protein IDs
dataSet1_notNormalized %>% 
  dplyr::filter(!is.na(Protein_id)) %>%
  dplyr:: arrange(Protein_id) -> dataSet1_notNormalized

# check for repeated genes
dataSet1_notNormalized %>% 
  dplyr::group_by(Protein_id) %>% 
  dplyr::summarize(n=n()) %>%
  dplyr::filter(!n==1) %>% 
  .$Protein_id %>% 
  as.vector(.)-> repeatedProtId1


dataSet1_notNormalized %>% 
  dplyr::filter(Protein_id %in% repeatedProtId1)

# it seems the data have repeateed proteins; I pick the ones with higher gene read

dataSet1_notNormalized %>%
  dplyr::mutate(glucose = as.numeric(as.character(glucose)),
                glycerol = as.numeric(as.character(glycerol)),
                Na_50_stress = as.numeric(as.character(Na_50_stress)),
                stationary = as.numeric(as.character(stationary)),
                late_stationary = as.numeric(as.character(late_stationary)))%>%
  na.omit(.)%>%
  dplyr::group_by(Protein_id) %>%
  dplyr::arrange(desc(glucose)) %>%
  dplyr::filter(row_number()==1) %>%
  dplyr::group_by() %>%
  dplyr::arrange(Protein_id)->dataSet1_notNormalized

# Check if everything is correct
dataSet1_notNormalized %>%
  dplyr::filter(gene_name %in% c("rpmE"))
# ***********************************************


# Making changes for dataSet2
# ***********************************************

# combine dictionary and datatSet2
dictionary %>%
  dplyr::select(gene_name, Protein_id) %>%
  dplyr::left_join(dataSet2_notNormalized, .) -> dataSet2_notNormalized

# get rid of not matching protein IDs
dataSet2_notNormalized %>% 
  dplyr::filter(!is.na(Protein_id)) %>%
  dplyr:: arrange(Protein_id) -> dataSet2_notNormalized

# check for repeated genes
dataSet2_notNormalized %>% 
  dplyr::group_by(Protein_id) %>% 
  dplyr::summarize(n=n()) %>%
  dplyr::filter(!n==1) %>% 
  .$Protein_id %>% 
  as.vector(.)-> repeatedProtId2

dataSet2_notNormalized %>% 
  dplyr::filter(Protein_id %in% repeatedProtId2)

# it seems the data have repeateed proteins; I pick the ones with higher gene read
dataSet2_notNormalized %>%
  dplyr::mutate(glucose = as.numeric(as.character(glucose)),
                glycerol = as.numeric(as.character(glycerol)),
                Na_50_stress = as.numeric(as.character(Na_50_stress)),
                stationary = as.numeric(as.character(stationary)),
                late_stationary = as.numeric(as.character(late_stationary)))%>%
  na.omit(.)%>%
  dplyr::group_by(Protein_id) %>%
  dplyr::arrange(desc(glucose)) %>%
  dplyr::filter(row_number()==1) %>%
  dplyr::group_by() %>%
  dplyr::arrange(Protein_id)->dataSet2_notNormalized

# Check if everything is correct
dataSet2_notNormalized %>%
  dplyr::filter(gene_name %in% c("bioD", "clpB"))


head(dataSet2_notNormalized)

summary(dataSet2_notNormalized)
# ***********************************************
```

Note: 
The `dataSet1_notNormalized` composed of only 1592 lines __which is few!!__
The `dataSet2_notNormalized` composed of only 1722 lines __which is few!!__


The next thing is to generate the _meta data_ data frame

with columns

* sampleNum (1:5)
* dataSet (externalS1: externamS5)
* experiment (external)
* carbonSource (glucose vs)
* growthPhase (exponential, stationary, late_stationary)
* Mg_mM_Levels (baseMg)
* Na_mM_Levels (baseNa, highNa)
* conditionInvestigated (Combination of all others)

and with row names equal to the columns of data matrix. 

```{r generate meta data frame and data matricies , echo=FALSE}

# generate meta data file
sampleNum = seq(1,5)
dataSet = c("glucose", "glycerol", "Na_50_stress", "stationary", "late_stationary")
experiment = rep("external",5)
carbonSource = c("glucose", "glycerol", "glucose", "glucose", "glucose")
growthPhase = c("exponential", "exponential", "exponential", "stationary", "late_stationary")
Mg_mM_Levels = rep("baseMg", 5)
Na_mM_Levels = c("baseNa", "baseNa", "highNa", "baseNa", "baseNa")
conditionInvestigated = c("Base Na_Base Mg_Glucose_Exponential","Base Na_BaseMg_Glycerol_Exponential", "High Na_Base Mg_Glucose_Exponential", "Base Na_Base Mg_Glucose_Stationary", "Base Na_Base Mg_Glucose_Late-Stationary")

meta_data = data.frame(sampleNum, 
                       dataSet, 
                       experiment, 
                       carbonSource, 
                       growthPhase,
                       Mg_mM_Levels,
                       Na_mM_Levels,
                       conditionInvestigated)

row.names(meta_data) <- c("glucose", 
                          "glycerol", 
                          "Na_50_stress", 
                          "stationary", 
                          "late_stationary")

print(meta_data)


# generate raw data matrix from DF for dataset 2
dataSet1_notNormalized %>%
  dplyr::group_by()%>%
  dplyr::select(glucose, glycerol, Na_50_stress, stationary, late_stationary)%>%
  as.matrix(.) -> rawData_matrix1
row.names(rawData_matrix1) <- dataSet1_notNormalized$Protein_id 

head(rawData_matrix1)


# generate raw data matrix from DF for dataset 2
dataSet2_notNormalized %>%
  dplyr::group_by()%>%
  dplyr::select(glucose, glycerol, Na_50_stress, stationary, late_stationary)%>%
  as.matrix(.) -> rawData_matrix2
row.names(rawData_matrix2) <- dataSet2_notNormalized$Protein_id 

head(rawData_matrix2)
```


Now we need to normalize the data. The process composed of 3 distinct steps:

* rounding the data (data composes of rounded integers, I do not need to do this step)
* DeSeq2 size factor normalization (with plus 1 size factor trick, to handle with "0" efficiently)
* DeSeq2 vst normalization

To do all that we need DESeq2 object

```{r apply DESeq Normalization with +1 matrix}
# For data set 1

# generate DeSeq2 P1 object from data matrix with trivial design formula
deseq_DataObj1_p1 <- DESeq2::DESeqDataSetFromMatrix(countData = rawData_matrix1+1, 
                                                    colData = meta_data, 
                                                    design = ~ 1) 
# calculate size factors
deseq_DataObj1_p1 <- DESeq2::estimateSizeFactors(deseq_DataObj1_p1)
sizeFactors1_p1 <- DESeq2::sizeFactors(deseq_DataObj1_p1)

# generate DeSeq2 object from data matrix with trivial design formula
deseq_DataObj1 <- DESeq2::DESeqDataSetFromMatrix(countData = rawData_matrix1, 
                                                 colData = meta_data, 
                                                 design = ~ 1) 

# Import size factors from P1 object
sizeFactors(deseq_DataObj1) <- sizeFactors1_p1


# For data set 2

# generate DeSeq2 P1 object from data matrix with trivial design formula
deseq_DataObj2_p1 <- DESeq2::DESeqDataSetFromMatrix(countData = rawData_matrix2+1, 
                                                    colData = meta_data, 
                                                    design = ~ 1) 
# calculate size factors
deseq_DataObj2_p1 <- DESeq2::estimateSizeFactors(deseq_DataObj2_p1)
sizeFactors2_p1 <- DESeq2::sizeFactors(deseq_DataObj2_p1)

# generate DeSeq2 object from data matrix with trivial design formula
deseq_DataObj2 <- DESeq2::DESeqDataSetFromMatrix(countData = rawData_matrix2, 
                                                 colData = meta_data, 
                                                 design = ~ 1) 

# Import size factors from P1 object
sizeFactors(deseq_DataObj2) <- sizeFactors2_p1
```


Than do the vst normalization (something similar to log normalization but reccommended by DESeq2) for datasets 1 & 2

```{r apply variance stabilizing transformation}
deseq_DataObj1 = DESeq2::varianceStabilizingTransformation(deseq_DataObj1)
deseq_DataObj2 = DESeq2::varianceStabilizingTransformation(deseq_DataObj2)
```

Finally Decompose The DESeq2 data into meta data and a raw data frame that can be used by machine learning algorithms.

```{r decompose DESeq2 obj}

# For data set 1
res_df1_ext<-as.data.frame(assay(deseq_DataObj1))
metaData1_ext<-as.data.frame(colData(deseq_DataObj1))

head(metaData1_ext)
head(res_df1_ext)

# For data set 2
res_df2_ext<-as.data.frame(assay(deseq_DataObj2))
metaData2_ext<-as.data.frame(colData(deseq_DataObj2))

head(metaData2_ext)
head(res_df2_ext)

# Save files
write.csv(x = metaData1_ext, file = "metaData1_ext.csv",row.names = TRUE)
write.csv(x = res_df1_ext, file = "res_df1_ext.csv",row.names = TRUE)

write.csv(x = metaData2_ext, file = "metaData2_ext.csv",row.names = TRUE)
write.csv(x = res_df2_ext, file = "res_df2_ext.csv",row.names = TRUE)
```

Two saved objects describe the cleaned and transformed data that is ready for machine learning applications. 

The machine learning part of the investigation can be found in part2